Synthesizing Verified Components for Cyber Assured Systems Engineering \thanks{DISTRIBUTION STATEMENT A. Approved for public release.}

### 2022-03-27 paper/agree-semantics.tex
wordsmithing on example; discussion of event ports tidied

The syntax of AGREE is
essentially that of quantifier-free first order predicate logic
supplemented with a few temporal operators. The terms (\emph{e}) are
arithmetic expressions built from variables ($v$) and numeric and
boolean literals ($c$), while formulas (\emph{b}) are built using
logical connectives from atomic formulas (\emph{a}) based on the
familiar comparison operators.
\[
\begin{array}{rcl}
e & ::= & v \mid c \mid e \;\set{+,*,/}\; e \\
a & ::= & e\; \set{=,<}\; e \\
b & ::= & v \mid c \mid a \mid \neg b
            \mid b \; \set{\land,\lor,\imp,\iff}\; b
\end{array}
\]
There is also a conditional, $\itelse{b}{(-)}{(-)}$, for both terms
and formulas. Lastly, there are temporal operators \emph{previously} $\konst{pre}(-)$,
\emph{followed-by} $(-) \to (-)$, and historically $\konst{Hist}(-)$.
These temporal operators are sufficient to define all other past-time operators \cite{monitor}.

The semantics of terms and formulas, as mentioned in the previous section, is in terms of \emph{streams of
values}. Values encompass at least booleans and numbers, but can be
readily extended to include records and arrays. A value stream is a
total function from time (natural numbers including 0) to values:
\[
 \konst{stream} = \mathbb{N}_0 \to \konst{value}
\]

Given an \emph{environment} $E : \konst{name} \mapsto \konst{stream}$
binding variable names to value streams, the semantics $\sem{-}^E_t$
of terms and formulas defines the meaning of compound syntax in terms
of the meaning of subexpressions. The value of a variable $v$ at time
$t$ is found by looking up the stream bound to $v$ in $E$ (call it
$s$) and returning $s_t$. Some clauses of the semantics follow,
omitting the temporal operators:
\[
\begin{array}{rcl}
\sem{v}^E_t & = & E(v)(t) \\
\sem{c}^E_t & = & c \\
\sem{e_1 + e_2}^E_t & = & \sem{e_1}^E_t + \sem{e_2}^E_t \\
   & \cdots & \\
\sem{b_1 \land b_2}^E_t & = & \sem{b_1}^E_t \land \sem{b_2}^E_t \\
   & \cdots & \\
\end{array}
\]

>>>>

The syntax of AGREE is essentially that of quantifier-free first order
predicate logic supplemented with a few temporal operators. The terms
(\emph{e}) are arithmetic expressions built from variables ($v$) and
numeric and boolean literals ($c$), while formulas (\emph{b}) are
built using logical connectives from atomic formulas (\emph{a}) based
on the familiar comparison operators.
\[
\begin{array}{rcl}
e & ::= & v \mid c \mid e \;\set{+,*,/}\; e \\
a & ::= & e\; \set{=,<}\; e \\
b & ::= & v \mid c \mid a \mid \neg b
            \mid b \; \set{\land,\lor,\imp,\iff}\; b
\end{array}
\]
There is also a conditional, $\itelse{b}{(-)}{(-)}$, for both terms
and formulas. To support port handling, there is a predicate
$\konst{event}(p)$, which tests if an event (or event-data) port $p$ is
signaled. Lastly, there are temporal operators \emph{previously}
$\konst{pre}(-)$, \emph{followed-by} $(-) \to (-)$, and historically
$\konst{Hist}(-)$.

The semantics of terms and formulas, as mentioned in the previous
section, is in terms of \emph{streams of values}. Values encompass at
least booleans and numbers, but can be readily extended to include
records and arrays. A value stream is a total function from time
(natural numbers including 0) to values:
\[
 \konst{stream} = \mathbb{N}_0 \to \konst{value}
\]
Given an \emph{environment} $E : \konst{name} \mapsto \konst{stream}$
binding variable names to value streams, the semantics $\sem{-}^E_t$
of terms and formulas defines the meaning of compound syntax in terms
of the meaning of subexpressions. The value of a variable $v$ at time
$t$ is found by looking up the stream bound to $v$ in $E$ (call it
$s$) and returning $s_t$. Some clauses of the semantics follow,
omitting the temporal operators:
\[
\begin{array}{rcl}
\sem{v}^E_t & = & E(v)(t) \\
\sem{c}^E_t & = & c \\
\sem{e_1 + e_2}^E_t & = & \sem{e_1}^E_t + \sem{e_2}^E_t \\
   & \cdots & \\
\sem{b_1 \land b_2}^E_t & = & \sem{b_1}^E_t \land \sem{b_2}^E_t \\
   & \cdots & \\
\end{array}
\]

### 2022-03-27 paper/example.tex
wordsmithing on example; discussion of event ports tidied

\figref{fig:example} is an AADL architectural model of a software
system (SW) for route planning and automated control of a UAV.  It is
loosely based on one of the case studies in
Section~\ref{sec:case-study}.  The source for the entire model is
found at \cite{repo}.

The system receives an automation request that is forwarded to a
third-party route planner (AI).  The route planner decides the flight
path of the UAV based on its current position and the requested task.
The waypoint manager (WM) receives the mission command as a set of
waypoints from the planner and starts the UAV flying the mission with an event on the \texttt{start} output port,
issuing waypoints to the UAV flight controller as the UAV location
changes.  The waypoint manager is a legacy component that cannot be
modified so it critically relies on assumptions about its input behavior to guarantee its intended output behavior.

The expected behavior of the SW system, and the components
implementing the system, are modeled with AGREE contract
specifications.  The contracts constrain input with assumptions and
state properties of output with guarantees if the assumptions are satisfied.  AGREE performs model
checking on this assume-guarantee system to hierarchically prove that
the SW component obeys all contract obligations under all possible finite input
streams.  The AGREE analysis, contract language, and contracts for SW
are discussed in greater detail in \secref{sec:agree}.

The initial AGREE contracts for the system in \figref{fig:example},
and its implementing components, both assume and guarantee the absence
of any malicious, or unspecified, component behavior.  More precisely,
the system contract assumes that all inputs are \emph{well-formed} and
there is never more than one automation request pending at a time.
The meaning of well-formed is given by user-written predicates for
each data type.  For example, a waypoint is well-formed if it falls
within bounds for latitude, longitude, and altitude.  The guarantees
for SW ensure that
\begin{compactitem}
\item a start includes a waypoint;
\item a start is within one cycle of an automation request and if not, then it persistently alerts;
\item new waypoints coincide with air vehicle location updates; and
\item all outputs are well-formed.
\end{compactitem}

The contracts for the sub-components assume their inputs are
well-formed, and they guarantee their outputs are well-formed.  The AI
contract guarantees it only responds to automation requests and always
in the same cycle.  The legacy WM contract guarantees the following if its input assumptions hold: 
\begin{compactitem}
  \item it generates a start from a response from the AI and always in the same cycle; 
  \item the start always comes with a waypoint; and 
  \item any further waypoints coincide with air vehicle location updates.  
\end{compactitem}
These initial specifications pass AGREE
verification meaning that the contract composition of the components
with the system satisfies all component input assumptions and system
output guarantees.

@@

\subsection{Code synthesis}
High-assurance components are automatically synthesized by SPLAT from
the AGREE specifications to equivalent programs in the CakeML
language.  The synthesis toolchain generates a proof that equates the
meaning of the AGREE specification to the behavior of the CakeML
program.  In other words, for any set of input streams that meet the
component's contract assumptions, the output streams produced from the
synthesized CakeML code exactly match those defined by the
high-assurance component's guarantees.  CakeML provides verified
compilation to binaries for several different platforms, meaning that
the resulting binaries exactly preserve the meaning of the original
CakeML code \cite{cakeml}.

>>>>

We now describe a simple software system, named SW, for route planning
and automated control of a UAV. A picture of the AADL architectural
model for SW is in \figref{fig:example}. SW is loosely based on one of
the case studies in Section~\ref{sec:case-study}.  The source for the
entire model is found at \cite{repo}.

From time to time, SW receives an \texttt{AutomationRequest} message,
forwarding it to a route planner (AI). The function of AI is to
compute the flight path (a list of waypoints) for the UAV, outputting
the resulting \emph{mission command} on \texttt{AutomationResponse}.
The waypoint manager (WM) receives the mission command from AI and
starts the UAV flying the mission by putting an \emph{event} on the
\texttt{Start} output port, continuing to issue waypoints to the UAV
flight controller on the \texttt{Waypoint} port as the UAV location
changes.\footnote{\konst{KLS}:What about Alert? When raised, is it
persistent, where is it computed?}  The AI component is third-party
software and WM is a legacy component.

%% that cannot be modified so it critically relies on
%% assumptions about its input behavior to guarantee its intended output
%% behavior.

The expected behavior of the SW system, and the components
implementing the system, is modeled with AGREE contract
specifications.  The contracts constrain input with assumptions and
state properties of output with guarantees.  AGREE performs model
checking on this assume-guarantee system to hierarchically prove that
the SW component obeys all contract obligations under all possible
finite input streams.  The AGREE analysis, contract language, and
contracts for SW are discussed in greater detail in
\secref{sec:agree}.

The AGREE contracts for the initial system in \figref{fig:example},
and its implementing components, both assume and guarantee the absence
of any malicious, or unspecified, component behavior.  More precisely,
the contract for SW assumes that its inputs are well-formed and that
there is never more than one automation request pending at a time.
(\emph{Well-formed} generally refers to a syntactic restriction on
data at a port. For example, a waypoint is well-formed if it falls
within bounds for latitude, longitude, and altitude.)  The guarantees
for SW ensure that
\begin{compactitem}
\item a start includes a waypoint;\footnote{\konst{KLS}: needs clarification; Start is just an event port.}
\item a start is within one cycle of an automation request and if not, then it persistently alerts;
\item new waypoints coincide with air vehicle location updates; and
\item all outputs are well-formed.
\end{compactitem}

The contracts for the sub-components assume their inputs are
well-formed, and they guarantee their outputs are well-formed.  The AI
contract guarantees it only responds to automation requests and always
in the same cycle.  The legacy WM contract guarantees the following if
its input assumptions hold:
\begin{compactitem}
  \item it generates a start from a response from the AI and always in the same cycle;
  \item the start always comes with a waypoint; and
  \item any further waypoints coincide with air vehicle location updates.
\end{compactitem}
These initial specifications pass AGREE verification, meaning that the
contract composition of the components with the system satisfies all
component input assumptions and system output guarantees.

@@

\subsection{Code synthesis}
High-assurance components are automatically synthesized by SPLAT from
the AGREE specifications to equivalent programs in the CakeML
language.
%% Too strong at the moment!
%The synthesis toolchain generates a proof that equates the
%% meaning of the AGREE specification to the behavior of the CakeML
%% program.
 In other words, for any set of input streams that meet the
 component's contract assumptions, the output streams produced from
 the synthesized CakeML code exactly match those defined by the
 high-assurance component's guarantees.  The CakeML compiler rovides
 verified compilation to binaries for several different platforms,
 meaning that the resulting binaries exactly preserve the meaning of
 the original CakeML code \cite{cakeml}.

### 2022-03-27 paper/verification-conditions.tex
Remove sentence on cut point in cycle
wordsmithing on example; discussion of event ports tidied

The AGREE specification language, and semantics, are inspired by the Lustre language \cite{10.1145/41625.41641}.
Lustre reasons over a synchronous data-flow model where the inputs and outputs of
components are streams.
Informally, streams assign values to expressions at each point in time.
AGREE contracts express relationships between input and output streams by assuming a set of allowed input streams and then guaranteeing a corresponding set of allowed output streams.
Output for streams not included in the assumed input is undefined.

When considering a system of components,
data flows through the components in dependency order, with inputs
being propagated to outputs through all contracts until they stabilize
(i.e., they cannot propagate further). Therefore, the subcomponent contracts, and
thus the top-level model, must be acyclic. 
Cyclic systems are made acyclic by breaking cycles with delays.
That is to say that input at the cut point reasons about previous input data rather than the current input data.
Once the data propagation
has stabilized, the model proceeds to the next input data in the input
streams. It is worth noting that the semantics do not model computation or communication
delay. The output of one contract is seen at the input of any
downstream contract in the same step of the input data stream.

>>>>

The AGREE specification language, and semantics, are inspired by the
Lustre language \cite{10.1145/41625.41641}. Lustre semantics are
synchronous data-flow where the inputs and outputs of components are
streams.  Informally, streams assign values to expressions at each
point in time.  AGREE contracts express relationships between input
and output streams by assuming a set of allowed input streams and then
guaranteeing a corresponding set of allowed output streams.

\subsection{Components}
A system implementation is described by its subcomponents plus
connections between their ports.  When considering a system of
components, data flows through the components in dependency order,
with inputs being propagated to outputs through all contracts until
they stabilize (i.e., they cannot propagate further). Therefore, the
subcomponent contracts, and thus the top-level model, must be acyclic.
Cyclic systems are made acyclic by breaking cycles with delays.  Once
the data propagation has stabilized, the model proceeds to the next
input data in the input streams. It is worth noting that the semantics
do not model computation or communication delay. The output of one
contract is seen at the input of any downstream contract in the same
step of the input data stream.

An component in AGREE is described relative to its \emph{ports}, a
concept inherited from AADL.  There are three types of AADL ports:
\emph{event} ports, \emph{data} ports, and \emph{event data} ports.
An event port has no data and only signals occurrence (or not) of the
event. A data port is a sampled port: the data can be read at anytime.
An event data port indicates when data is present with an event.

\subsection{Contracts and verification conditions}
